Binocular integration of chromatic and luminance signals

### 2024-04-15 manuscript.tex
Switched to using normalized contrast values

\begin{document}
\maketitle
\(^1\)Department of Psychology, University of York, UK

\(^2\)York Biomedical Research Institute, University of York, UK

\(^3\)School of Psychology, University of Nottingham, UK

\(^4\)School of Philosophy, Psychology and Language Sciences, University
of Edinburgh, UK

\(^*\)Corresponding author, email: daniel.baker@york.ac.uk

@@

Much progress has been made in understanding how the brain combines
signals from the two eyes. However most of this work has involved
achromatic (black and white) stimuli, and it is not clear if the same
processes apply in colour-sensitive pathways. In our first experiment,
we measured contrast discrimination (`dipper') functions for four key
ocular configurations (monocular, binocular, half-binocular and
dichoptic), for achromatic, isoluminant red/green and isoluminant
blue/yellow sine-wave grating stimuli. We find a similar pattern of
results across stimuli, implying equivalently strong interocular
suppression within each pathway. Our second experiment measured
dichoptic masking within and between pathways using the method of
constant stimuli. Masking was strongest within-pathway, and weakest
between chromatic and achromatic mechanisms. Finally, we repeated the
dipper experiment using temporal luminance modulations, which produced
slightly weaker interocular suppression than for spatially modulated
stimuli. We interpret our results in the context of a contemporary
two-stage model of binocular contrast gain control, implemented here
using a hierarchical Bayesian framework. Posterior distributions of the
weight of interocular suppression overlapped with a value of 1 for all
dipper data sets, and the model captured well the pattern of threshold
and slope values from all three experiments.

@@

The process by which the brain combines independent inputs is of
fundamental importance for understanding sensory perception. Binocular
vision is a useful test-case for determining the general principles
involved in neural signal combination, as our brains typically combine
the inputs from the left and right eyes to provide binocular single
vision. In recent years our understanding has been facilitated by the
development of binocular gain control models that provide a framework to
interpret empirical data from multiple experimental paradigms and
techniques, including psychophysics (Meese et al., 2006), EEG (Baker \&
Wade, 2017), fMRI (Moradi \& Heeger, 2009) and pupillometry (Segala et
al., 2023). However, the majority of this work has used achromatic
(black and white) stimuli; we know comparatively little about how
chromatic signals are combined binocularly, or about how signals in
different ocular and chromatic channels interact. In this study we use
psychophysical detection and discrimination paradigms to explore
binocular interactions within and between the chromatic and achromatic
pathways.

@@

The detailed pattern of thresholds across these four conditions is
complex, and for achromatic stimuli has several distinctive features
that have been replicated in multiple studies. At low pedestal
contrasts, the binocular condition yields lower thresholds than the
monocular condition; this result is attributed to physiological
binocular summation by neurons responsive to signals from both eyes
(Baker et al., 2018; Campbell \& Green, 1965). However at high pedestal
contrasts the `handle' regions of the dipper functions for these
conditions converge: a consequence of interocular suppression
compensating for the increased excitation during binocular stimulation
(Legge, 1984; Maehara \& Goryo, 2005). The half-binocular condition
avoids confounding the number of eyes seeing the target with the number
of eyes seeing the pedestal (Meese et al., 2006). The pedestal is always
binocular in this condition, whereas the target increment is monocular,
and thresholds are consistently higher than in the binocular condition
across the full range of pedestal contrasts. This demonstrates that
binocular summation occurs across the full contrast range, when the
pedestal ocularity is appropriately controlled. Finally, the dichoptic
condition produces extremely strong masking of the target, such that
when the pedestal is visible, the target must equal or exceed its
contrast in order to be detectable (Baker \& Meese, 2007; Legge, 1979;
Maehara \& Goryo, 2005). The characteristic pattern of dipper functions
is well-described by the gain control model of Meese et al. (2006) for
achromatic stimuli.

At the output of the human retina, cone responses are split into three
distinct pathways. The sum of long- and medium-wavelength cone outputs
(L+M) transmits luminance information, and is likely responsible for the
binocular combination effects previously studied using achromatic
stimuli (see above). The difference of long- and medium-wavelength cone
outputs (L-M) is responsive to chromatic stimuli modulating along a
red/green axis in colour space. Finally the short wavelength cone
outputs (S-(L+M)) code chromatic stimuli modulating along a blue/yellow
axis. There has not yet been a detailed investigation of binocular
contrast discrimination in either of these chromatic pathways, however
there is reason to believe they may differ from the achromatic pathway.
At detection threshold, binocular summation is greater for chromatic
versus achromatic stimuli (Simmons, 2005), implying a more linear
initial stage of processing. For cross-orientation masking, there are
differences in the magnitude of masking between chromatic (red/green)
and achromatic stimuli (Kim et al., 2013; Medina \& Mullen, 2009), as
well as differences in their temporal dynamics (Kim \& Mullen, 2015).
There are also interactions between chromatic and achromatic pathways
both within (Chen et al., 2000) and between (Kingdom \& Libenson, 2015;
Mullen et al., 2014) the eyes, yet these have not been fully explored
for arrangements where the target and mask have the same orientation.
Finally, the neurophysiological underpinnings of colour vision are
distinct from those of the achromatic system. In primary visual cortex
(V1), chromatic signals are processed in `blob' regions that are
revealed by cytochrome oxidase staining (Horton \& Hubel, 1981). The
blob regions appear to be largely monocular (Livingstone \& Hubel,
1984), suggesting that binocular combination for chromatic stimuli might
occur later than for achromatic stimuli, and perhaps be subject to
different constraints.

A further property of `blob' regions is that they are less strongly
orientation-tuned than other regions of V1 (Horton \& Hubel, 1981;
Livingstone \& Hubel, 1984), and biased towards low spatial frequencies
(Edwards et al., 1995; Tootell et al., 1988). Our recent work (Segala et
al., 2023) has investigated binocular combination for flickering discs
of luminance, which are DC-balanced across time (i.e.~the time-averaged
luminance is equal to the background). Steady-state EEG responses from
early visual cortex and psychophysical contrast matching data were both
consistent with weak interocular suppression when using this stimulus
arrangement. This is very different from the `ocularity invariance'
(where binocular and monocular stimuli appear equal) that is
well-established when using DC-balanced periodic stimuli such as
sine-wave gratings, and implies strong interocular suppression (Baker \&
Wade, 2017; Meese et al., 2006; Moradi \& Heeger, 2009). There is
additional psychophysical evidence from matching studies that binocular
combination can be close to linear for luminance increments (Anstis \&
Ho, 1998; Levelt, 1965), particularly against a dark background (Baker
et al., 2012). Both structural and functional data therefore imply that
binocular combination may differ between spatial and temporal contrast.

The main aim of the present study was to characterise binocular signal
combination for chromatic stimuli, and for temporal modulations of
luminance. We also aimed to investigate interocular suppression between
chromatic and achromatic pathways. We therefore preregistered a series
of psychophysical experiments (see: \url{https://osf.io/3vdga/}). In
Experiment 1 we replicate the four key pedestal masking conditions of
Meese et al. (2006) described above for achromatic grating stimuli, and
extend this to both red/green and blue/yellow isoluminant chromatic
stimuli. In Experiment 2 we explore dichoptic masking within and between
these stimuli. Experiment 3 repeats the achromatic condition from the
first experiment, but using a temporally modulated disc rather than
sine-wave gratings. We take a Bayesian approach to data analysis and
modelling; by fitting a hierarchical version of the two-stage gain
control model (Meese et al., 2006) we compare posterior parameter
distributions to understand how model parameters such as the weight of
interocular suppression vary across visual pathways.

@@

In Experiments 1 and 2, the stimuli were horizontal sinusoidal gratings
with a spatial frequency of 1c/deg (see examples in
Figure~\ref{fig-examplestim}). The gratings were windowed by a raised
cosine envelope with a diameter of 3 degrees. Spatial phase, relative to
a central fixation cross, was randomised on each trial across the four
cardinal phases. In the achromatic conditions, the sine-wave modulated
all three monitor colour channels equally (red, green and blue). In the
L-M condition, we generated isoluminant stimuli for each participant
(see Procedures) designed to maximise contrast between L and M cones,
whilst keeping S cone activity constant. In the S-(L+M) condition, the
isoluminant stimuli maximised S-cone contrast. Stimuli were converted
from cone space to monitor RGB coordinates using the monitor spectral
readings and the Stockman-Sharpe 2 degree cone fundamentals (Stockman \&
Sharpe, 2000). The stimuli in Experiment 3 were temporal modulations of
luminance applied to a disc made using the same raised cosine envelope
as described above, but with no further spatial modulation. The stimuli
counterphase flickered sinusoidally at 4Hz (see examples in the lower
portion of Figure~\ref{fig-examplestim}). In all experiments, we
displayed a binocular fusion lock, consisting of three concentric rings
of small square elements with random colour. A black central fixation
cross was also displayed throughout.

@@

All stimuli were presented on an Iiyama VisionMaster Pro 510 CRT
monitor, with a refresh rate of 100Hz, and a resolution of 1024 x 768
pixels. The display was driven by a ViSaGe MkII stimulus generator
(Cambridge Research Systems Ltd., Kent, UK) running in 42-bit colour
mode (14 bits per colour channel). We presented stimuli to the left and
right eyes independently using a four-mirror stereoscope with
front-silvered mirrors. The display was luminance calibrated using a
ColourCal photometer (Cambridge Research Systems), and gamma corrected
by fitting a four-parameter gamma function to the output of each CRT
gun. The maximum luminance was 87 cd/m\(^2\). We also measured the
spectral output of each phosphor using a Jaz spectroradiometer (Ocean
Insight, Florida), and used these measurements to convert between LMS
(cone) space and the monitor RGB coordinates.

For convenience, we express stimulus contrast as a percentage of the
maximum possible contrast that could be displayed on our system. For
achromatic stimuli, the maximum contrast is 1, so this is equivalent to
the standard Michelson contrast expressed as a percentage. For the
isoluminant chromatic stimuli, the maximum displayable L-M (red/green)
cone contrast was 0.1, and the maximum displayable S-(L+M) (blue/yellow)
cone contrast was 0.88. An L-M threshold of 50\% therefore corresponds
to a cone contrast of \(0.5\times0.1 = 0.05\), and an S-(L+M) threshold
of 50\% corresponds to a cone contrast of \(0.5\times0.88 = 0.44\). This
means that the threshold values reported throughout can be converted to
cone contrast by a straightforward multiplicative transform.

@@

Before beginning primary data collection, each participant in
Experiments 1 and 2 completed an isoluminance adjustment task. Grating
stimuli were presented that counterphase flickered at 5Hz, defined about
either the L-M or S-(L+M) plane in cone space. Participants used a
trackball to dynamically adjust the colour angle of the stimulus to
minimise the percept of flicker. Each participant completed ten such
trials for each colour plane, and the average angle across repetition
was taken as the isoluminant point, and used to generate stimuli for the
main experiment for that participant. Settings were very similar across
participants for the S-(L+M) direction, and somewhat more heterogeneous
for the L-M direction (see Figure~\ref{fig-isofig}).

@@

In Experiment 1, participants completed a two-interval-forced-choice
(2IFC) contrast discrimination task. Stimuli were presented for 200ms,
with an interstimulus interval of 400ms. Each interval was indicated by
an auditory beep, and participants made their responses using a
two-button trackball. Correct responses were followed by a high pitched
tone, and incorrect responses by a low pitched tone. Each block of the
experiment tested a single pedestal contrast level, and lasted around 12
minutes. On each trial the target contrast level was determined by a
3-down-1-up staircase procedure. There were 8 interleaved staircases in
total; four stimulus arrangements (see Figure~\ref{fig-exampledips})
combined factorially with two target eye assignments. Each pedestal
contrast was repeated 3 times by each participant, and the block order
was randomised. The experiment lasted around 4 hours per participant for
each chromatic condition, and took place over the course of several
weeks. In total, the experiment consisted of 67978 trials (pooled across
participants).

In Experiment 2, participants completed a 2IFC dichoptic masking task.
The stimuli and trial protocol were the same as for Experiment 1, except
that the target contrast was chosen from a set of 10 possible values,
determined in advance based on the data of Experiment 1. There were 12
possible conditions: baseline detection thresholds for achromatic,
red/green and blue/yellow stimuli, and the nine possible factorial
pairings obtained by assigning these conditions to be target and
dichoptic mask stimuli. Mask contrasts were chosen to be approximately
16 times their (monocular) detection threshold, based on the data from
Experiment 1. As a percentage of the maximum displayable contrast, these
were 16\% for achromatic stimuli, and 64\% for both of the chromatic
stimuli. Each block of the experiment tested a single condition, and
consisted of 200 trials. A high contrast example of the target stimulus
was displayed at the foot of the screen throughout, so that there was no
ambiguity about the target identity on a given block. Participants
completed 10 repetitions of each condition (120 blocks of
\textasciitilde6 minutes each), lasting around 12 hours, for a total of
72000 trials (pooled across participants).

@@

The two stage model of Meese et al. (2006) was fit to the threshold data
from Experiments 1 and 3 using a simplex algorithm to minimise the error
between the model and data. The model is defined by a series of
equations:

@@

where \(C_L\) and \(C_R\) are the contrasts displayed to the left and
right eyes, and \(m\), \(S\), \(\omega\), \(p\), \(q\) and \(Z\) are
free parameters in the model. A further free parameter, \(k\),
represents additive internal noise, and is used to convert the model
outputs to either d-prime or threshold values (note that in the original
model specification (Meese et al., 2006) this parameter was called
\(\sigma\), but we use the \(k\) symbol here to avoid confusion with the
standard deviation of the cumulative Gaussian used when fitting the
psychometric functions to estimate thresholds). Thresholds are defined
by iteratively adjusting the target contrast until the following
equality is satisfied:

@@

We also implemented a Bayesian hierarchical version of the model using
the Stan language (Carpenter et al., 2017). This used a binomial
generator function to model the proportion correct data at each target
level, and was fit simultaneously to all participants for a given
experiment, but separately for each chromatic condition of Experiment 1,
and the flickering disc data from Experiment 3 (i.e.~four fits in total,
as for the simplex fitting). Prior distributions were Gaussian, with
means determined from published values (see first row of
Table~\ref{tbl-parametertable}). This modelling primarily focuses on
examining posterior parameter distributions, rather than a model
comparison approach. We generated over 1 million posterior samples for
the model, and retained 10\% of them for plotting.

@@

where \emph{AC} represents the achromatic contrast, \emph{RG} represents
the red/green contrast, \emph{BY} represents the blue/yellow contrast,
and \(\omega_A\), \(\omega_R\) and \(\omega_B\) are the accompanying
weights of interocular suppression. There is an equivalent expression
for the right eye, and for each of the two isoluminant chromatic
pathways. To simplify the model, and avoid free parameters that are
poorly constrained by the data, we fixed several parameters at the
average values from the fits from Experiment 1
(Table~\ref{tbl-parametertable}), such that \(p=7.82\), \(q=6.4\),
\(m=1.21\), \(S=1.03\) and \(k=0.23\). We additionally fixed the
within-mechanism weight of interocular suppression (\(\omega\)) at 1.
This left 9 free parameters: a \(Z\) parameter for each mechanism, and
six cross-mechanism weights of interocular suppression. These parameters
were again estimated within a Bayesian hierarchical framework, using the
trial-wise data from Experiment 2. Note that the model as specified does
not currently include monocular suppression between different pathways,
as we did not collect any data for these conditions. Previous work (Chen
et al., 2000) has measured such interactions, and they could in
principle be incorporated into the denominator of either stage 1 or
stage 2 in the model.

@@

\caption{\label{fig-dipperfig}Dipper functions and psychometric slopes
from Experiment 1, averaged across three participants. Panels (a-c) show
threshold data, and panels (d-f) represent the slope of the psychometric
function expressed in Weibull \(\beta\) units. Error bars give ±1SE
across participants. Note that contrast values are expressed as a
percentage of the maximum displayable contrast (see Procedures for
details). Curves in panels (a-c) show the best fitting models, optimized
using a simplex algorithm, and RMSE values give the root mean squared
errors of the fits.}

@@

Following Meese et al. (2006), we also inspected the slope of the
psychometric function for each condition (see
Figure~\ref{fig-dipperfig}d-f). At detection threshold, slopes were
relatively steep for all stimuli, with values around \(\beta=4\). As
pedestal contrasts increased, slopes linearized and reduced to around
\(\beta=1.3\) (Foley \& Legge, 1981; Meese et al., 2006), and remained
shallow at high pedestal contrasts. The exception to this was the
dichoptic condition, where slopes became extremely steep at high
dichoptic mask contrasts, consistent with previous observations (Baker
et al., 2013; Meese et al., 2006). This was clear for all three data
sets, with slope values in the range \(4 < \beta < 8\). However, we
observe that slope estimates are more variable than threshold estimates
(note the large error bars), particularly when using adaptive
staircases, which deploy the majority of trials close to threshold. Our
second experiment therefore investigated the slope of the psychometric
function in more detail for dichoptic masking using the method of
constant stimuli, as well as exploring dichoptic interactions between
chromatic channels.

@@

\caption{\label{fig-MCSfig}Summary of data from Experiment 2. Panels
(a-c) show psychometric functions for each condition, pooled across
participants (600 trials per target contrast level). Panels (d-f) show
threshold and slope estimates for individual participants (points) and
the boundary of the posterior density estimates for fits to the pooled
data (ellipses). Panel (g) shows the average threshold elevation factor
for each combination of target and mask stimulus. Panel (h) shows the
geometric mean psychometric slope value for each masking condition in
Weibull \(\beta\) units.}

@@

Threshold elevation was greatest when the target and mask had the same
chromaticity - notice that the psychometric function is shifted furthest
to the right for the achromatic target with an achromatic mask (white
and black circles in Figure~\ref{fig-MCSfig}a), for the red/green target
with a red/green mask (red and green circles in
Figure~\ref{fig-MCSfig}b), and for the blue/yellow target with a
blue/yellow mask (blue and yellow circles in Figure~\ref{fig-MCSfig}c).
Masking was weakest between achromatic masks/targets and chromatic
masks/targets. Finally there was an intermediate level of masking
between red/green and blue/yellow stimuli. This is summarised in
Figure~\ref{fig-MCSfig}g, which represents threshold elevation for each
combination of target and mask chromaticity. Note that the positive
diagonal exhibits the highest values, and represents threshold elevation
between targets and masks of the same chromaticity.

>>>>

\begin{document}
\maketitle
\(^1\)Department of Psychology, University of York, UK\\
\(^2\)York Biomedical Research Institute, University of York, UK\\
\(^3\)School of Psychology, University of Nottingham, UK\\
\(^4\)School of Philosophy, Psychology and Language Sciences, University
of Edinburgh, UK\\
\(^*\)Corresponding author, email: daniel.baker@york.ac.uk

@@

Much progress has been made in understanding how the brain combines
signals from the two eyes. However, most of this work has involved
achromatic (black and white) stimuli, and it is not clear if the same
processes apply in colour-sensitive pathways. In our first experiment,
we measured contrast discrimination (`dipper') functions for four key
ocular configurations (monocular, binocular, half-binocular and
dichoptic), for achromatic, isoluminant red/green and isoluminant
blue/yellow sine-wave grating stimuli. We find a similar pattern of
results across stimuli, implying equivalently strong interocular
suppression within each pathway. Our second experiment measured
dichoptic masking within and between pathways using the method of
constant stimuli. Masking was strongest within-pathway, and weakest
between chromatic and achromatic mechanisms. Finally, we repeated the
dipper experiment using temporal luminance modulations, which produced
slightly weaker interocular suppression than for spatially modulated
stimuli. We interpret our results in the context of a contemporary
two-stage model of binocular contrast gain control, implemented here
using a hierarchical Bayesian framework. Posterior distributions of the
weight of interocular suppression overlapped with a value of 1 for all
dipper data sets, and the model captured well the pattern of threshold
and slope values from all three experiments.

@@

The process by which the brain combines independent inputs is of
fundamental importance for understanding sensory perception. Binocular
vision is a useful test-case for determining the general principles
involved in neural signal combination, as our brains typically combine
the inputs from the left and right eyes to provide binocular single
vision. In recent years our understanding has been facilitated by the
development of binocular gain control models that provide a framework to
interpret empirical data from multiple experimental paradigms and
techniques, including psychophysics (Meese et al., 2006), EEG (Baker \&
Wade, 2017), fMRI (Moradi \& Heeger, 2009) and pupillometry (Segala et
al., 2023). However, most of this work has used achromatic (black and
white) stimuli; we know comparatively little about how chromatic signals
are combined binocularly, or about how signals in different ocular and
chromatic channels interact. In this study we use psychophysical
detection and discrimination paradigms to explore binocular interactions
within and between the chromatic and achromatic pathways.

@@

The detailed pattern of thresholds across these four conditions is
complex, and for achromatic stimuli has several distinctive features
that have been replicated in multiple studies. At low pedestal
contrasts, the binocular condition yields lower thresholds than the
monocular condition; this result is attributed to physiological
binocular summation by neurons responsive to signals from both eyes
(Baker et al., 2018; Campbell \& Green, 1965). However, at high pedestal
contrasts the `handle' regions of the dipper functions for these
conditions converge: a consequence of interocular suppression
compensating for the increased excitation during binocular stimulation
(Legge, 1984; Maehara \& Goryo, 2005). The half-binocular condition
avoids confounding the number of eyes seeing the target with the number
of eyes seeing the pedestal (Meese et al., 2006). The pedestal is always
binocular in this condition, whereas the target increment is monocular,
and thresholds are consistently higher than in the binocular condition
across the full range of pedestal contrasts. This demonstrates that
binocular summation occurs across the full contrast range, when the
pedestal ocularity is appropriately controlled. Finally, the dichoptic
condition produces extremely strong masking of the target, such that
when the pedestal is visible, the target must equal or exceed its
contrast to be detectable (Baker \& Meese, 2007; Legge, 1979; Maehara \&
Goryo, 2005). The characteristic pattern of dipper functions is
well-described by the gain control model of Meese et al. (2006) for
achromatic stimuli.

At the output of the human retina, cone responses are split into three
distinct pathways. The sum of long- and medium-wavelength cone outputs
(L+M) transmits luminance information (see Sharpe et al., 2011), and is
likely responsible for the binocular combination effects previously
studied using achromatic stimuli (see above). The difference of long-
and medium-wavelength cone outputs (L-M) is responsive to chromatic
stimuli modulating along a red/green axis in colour space. Finally the
opponent (S-(L+M)) channel codes chromatic stimuli modulating along a
blue/yellow axis. At isoluminance, where (L+M) is held constant, this
channel is driven purely by the S-cone outputs.

There has not yet been a detailed investigation of binocular contrast
discrimination in either of these chromatic pathways, however there is
reason to believe they may differ from the achromatic pathway. At
detection threshold, binocular summation is greater for chromatic versus
achromatic stimuli (Simmons, 2005), implying a more linear initial stage
of processing. For cross-orientation masking, there are differences in
the magnitude of masking between chromatic (red/green) and achromatic
stimuli (Kim et al., 2013; Medina \& Mullen, 2009), as well as
differences in their temporal dynamics (Kim \& Mullen, 2015). There are
also interactions between chromatic and achromatic pathways both within
(Chen et al., 2000) and between (Kingdom \& Libenson, 2015; Mullen et
al., 2014) the eyes, yet these have not been fully explored for
arrangements where the target and mask have the same orientation.
Finally, the neurophysiological underpinnings of colour vision are
distinct from those of the achromatic system. In primary visual cortex
(V1), chromatic signals are processed in `blob' regions that are
revealed by cytochrome oxidase staining (Horton \& Hubel, 1981). The
blob regions appear to be largely monocular (Livingstone \& Hubel,
1984), suggesting that binocular combination for chromatic stimuli might
occur later than for achromatic stimuli, and perhaps be subject to
different constraints.

A further property of `blob' regions is that they are less strongly
orientation-tuned than other regions of V1 (Horton \& Hubel, 1981;
Livingstone \& Hubel, 1984), and biased towards low spatial frequencies
(Edwards et al., 1995; Tootell et al., 1988; though see Economides et
al., 2011). Our recent work (Segala et al., 2023) has investigated
binocular combination for flickering discs of luminance, which are
DC-balanced across time (i.e.~the time-averaged luminance is equal to
the background). Steady-state EEG responses from early visual cortex and
psychophysical contrast matching data were both consistent with weak
interocular suppression when using this stimulus arrangement. This is
very different from the `ocularity invariance' (where binocular and
monocular stimuli appear equal) that is well-established when using
DC-balanced periodic stimuli such as sine-wave gratings, and implies
strong interocular suppression (Baker \& Wade, 2017; Meese et al., 2006;
Moradi \& Heeger, 2009). There is additional psychophysical evidence
from matching studies that binocular combination can be close to linear
for luminance increments (Anstis \& Ho, 1998; Levelt, 1965),
particularly against a dark background (Baker et al., 2012). Both
structural and functional data therefore imply that binocular
combination may differ between spatial and temporal contrast.

The main aim of the present study was to characterise binocular signal
combination for chromatic stimuli, and for temporal modulations of
luminance. We also aimed to investigate interocular suppression between
chromatic and achromatic pathways. We therefore preregistered a series
of psychophysical experiments (see: \url{https://osf.io/3vdga/}). In
Experiment 1 we replicate the four key pedestal masking conditions of
Meese et al. (2006) described above for achromatic grating stimuli and
extend this to both red/green and blue/yellow isoluminant chromatic
stimuli. In Experiment 2 we explore dichoptic masking within and between
these stimuli. Experiment 3 repeats the achromatic condition from the
first experiment but using a temporally modulated disc rather than
sine-wave gratings. We take a Bayesian approach to data analysis and
modelling; by fitting a hierarchical version of the two-stage gain
control model (Meese et al., 2006) we compare posterior parameter
distributions to understand how model parameters such as the weight of
interocular suppression vary across visual pathways.

@@

In Experiments 1 and 2, the stimuli were horizontal sinusoidal gratings
with a spatial frequency of 1c/deg (see examples in
Figure~\ref{fig-examplestim}). The gratings were windowed by a raised
cosine envelope with a diameter of 3 degrees. Spatial phase, relative to
a central fixation cross, was randomised on each trial across the four
cardinal phases. In the achromatic conditions, the sine-wave modulated
all three monitor colour channels equally (red, green and blue). In the
L-M condition, we generated isoluminant stimuli for each participant
(see Procedures) designed to maximise contrast between L and M cones,
whilst keeping S cone activity constant. In the S-(L+M) condition, the
isoluminant stimuli maximised S-cone contrast. Stimuli were converted
from cone space to monitor RGB coordinates using the monitor spectral
readings and the Stockman-Sharpe 2 degree cone fundamentals (Stockman \&
Sharpe, 2000). The maximum displayable cone contrasts on our system were
0.1 for L-M, and 0.88 for S-(L+M). The stimuli in Experiment 3 were
temporal modulations of luminance applied to a disc made using the same
raised cosine envelope as described above, but with no further spatial
modulation. The stimuli counterphase flickered sinusoidally at 4Hz (see
examples in the lower portion of Figure~\ref{fig-examplestim}). In all
experiments, we displayed a binocular fusion lock, consisting of three
concentric rings of small square elements with random colour. A black
central fixation cross was also displayed throughout.

@@

All stimuli were presented on an Iiyama VisionMaster Pro 510 CRT
monitor, with a refresh rate of 100Hz, and a resolution of 1024 x 768
pixels. The display was driven by a ViSaGe MkII stimulus generator
(Cambridge Research Systems Ltd., Kent, UK) running in 42-bit colour
mode (14 bits per colour channel). We presented stimuli to the left and
right eyes independently using a four-mirror stereoscope with
front-silvered mirrors. The display was luminance calibrated using a
ColourCal photometer (Cambridge Research Systems), and gamma corrected
by fitting a four-parameter gamma function to the output of each CRT
gun. The maximum luminance was 87 cd/m\(^2\). We also measured the
spectral output of each phosphor using a Jaz spectroradiometer (Ocean
Insight, Florida), and used these measurements to convert between LMS
(cone) space and the monitor RGB coordinates. We express contrast values
in normalized units, relative to the monocular threshold for each
stimulus type.

@@

Before beginning primary data collection, each participant in
Experiments 1 and 2 completed an isoluminance adjustment task. Grating
stimuli were presented that counterphase flickered at 5Hz, with contrast
variations defined about either the L-M or S-(L+M) plane in cone space.
Participants used a trackball to dynamically adjust the colour angle of
the stimulus to minimise the percept of flicker. Each participant
completed ten such trials for each colour plane, and the average angle
across repetition was taken as the isoluminant point, and used to
generate stimuli for the main experiment for that participant. Settings
were very similar across participants for the S-(L+M) direction, and
somewhat more heterogeneous for the L-M direction (see
Figure~\ref{fig-isofig}).

@@

In Experiment 1, participants completed a two-interval-forced-choice
(2IFC) contrast discrimination task. Stimuli were presented for 200ms,
with an interstimulus interval of 400ms. Each interval was indicated by
an auditory beep, and participants made their responses using a
two-button trackball. Correct responses were followed by a high-pitched
tone, and incorrect responses by a low-pitched tone. Each block of the
experiment tested a single pedestal contrast level and lasted around 12
minutes. On each trial the target contrast level was determined by a
3-down-1-up staircase procedure. There were 8 interleaved staircases in
total; four stimulus arrangements (see Figure~\ref{fig-exampledips})
combined factorially with two target eye assignments. Each pedestal
contrast was repeated 3 times by each participant, and the block order
was randomised. The experiment lasted around 4 hours per participant for
each chromatic condition, and took place over the course of several
weeks. In total, the experiment consisted of 67978 trials (pooled across
participants).

In Experiment 2, participants completed a 2IFC dichoptic masking task.
The stimuli and trial protocol were the same as for Experiment 1, except
that the target contrast was chosen from a set of 10 possible values,
determined in advance based on the data of Experiment 1. There were 12
possible conditions: baseline detection thresholds for achromatic,
red/green and blue/yellow stimuli, and the nine possible factorial
pairings obtained by assigning these conditions to be target and
dichoptic mask stimuli. Mask contrasts were chosen to be approximately
16 times their (monocular) detection threshold, based on the data from
Experiment 1. The contrasts were 16\% Michelson contrast for the
achromatic stimuli, 6.4\% cone contrast for L-M and 56\% cone contrast
for S-(L+M). Each block of the experiment tested a single condition, and
consisted of 200 trials. A high contrast example of the target stimulus
was displayed at the foot of the screen throughout, so that there was no
ambiguity about the target identity on a given block. Participants
completed 10 repetitions of each condition (120 blocks of
\textasciitilde6 minutes each), lasting around 12 hours, for a total of
72000 trials (pooled across participants).

@@

The two-stage model of Meese et al. (2006) was fit to the threshold data
from Experiments 1 and 3 using a simplex algorithm to minimise the error
between the model and data. We normalized the thresholds and pedestal
contrasts to the appropriate monocular threshold for each stimulus type.
The model is defined by a series of equations:

@@

where \(C_L\) and \(C_R\) are the (normalized) contrasts displayed to
the left and right eyes, and \(m\), \(S\), \(\omega\), \(p\), \(q\) and
\(Z\) are free parameters in the model. A further free parameter, \(k\),
represents additive internal noise, and is used to convert the model
outputs to either d-prime or threshold values (note that in the original
model specification (Meese et al., 2006) this parameter was called
\(\sigma\), but we use the \(k\) symbol here to avoid confusion with the
standard deviation of the cumulative Gaussian used when fitting the
psychometric functions to estimate thresholds). Thresholds are defined
by iteratively adjusting the target contrast until the following
equality is satisfied:

@@

We also implemented a Bayesian hierarchical version of the model using
the Stan language (Carpenter et al., 2017). This used a binomial
generator function to model the proportion correct data at each target
level, and was fit simultaneously to all participants for a given
experiment, but separately for each chromatic condition of Experiment 1,
and the flickering disc data from Experiment 3 (i.e.~four fits in total,
as for the simplex fitting). Prior distributions for the parameters
\emph{p}, \emph{q}, \emph{m} and \(\omega\) were Gaussian, with means
determined from published values (see first row of
Table~\ref{tbl-parametertable}). Priors for parameters \emph{S},
\emph{Z} and \emph{k} were uniform. This modelling primarily focuses on
examining posterior parameter distributions, rather than a model
comparison approach. We generated over 1 million posterior samples for
the model, and retained 10\% of them for plotting.

@@

where \emph{AC} represents the achromatic contrast, \emph{RG} represents
the red/green contrast, \emph{BY} represents the blue/yellow contrast,
and \(\omega_A\), \(\omega_R\) and \(\omega_B\) are the accompanying
weights of interocular suppression. There is an equivalent expression
for the right eye, and for each of the two isoluminant chromatic
pathways. To simplify the model, and avoid free parameters that are
poorly constrained by the data, we fixed several parameters (\emph{p},
\emph{q}, \emph{m}, \emph{S}, \emph{k}, and \(\omega\) for the
within-pathway suppression) at the values from the fits from Experiment
1 (Table~\ref{tbl-parametertable}, lower rows). This left 9 free
parameters: a \(Z\) parameter for each mechanism, and six
cross-mechanism weights of interocular suppression. These parameters
were again estimated within a Bayesian hierarchical framework, using the
trial-wise data from Experiment 2. Note that the model as specified does
not currently include monocular suppression between different pathways,
as we did not collect any data for these conditions. Previous work (Chen
et al., 2000) has measured such interactions, and they could in
principle be incorporated into the denominator of either stage 1 or
stage 2 in the model.

@@

\caption{\label{fig-dipperfig}Dipper functions and psychometric slopes
from Experiment 1, averaged across three participants. Panels (a-c) show
threshold data, and panels (d-f) represent the slope of the psychometric
function expressed in Weibull \(\beta\) units. Error bars give ±1SE
across participants. Note that all contrast values for both thresholds
and pedestals are normalized to the monocular detection threshold in
each panel. Curves in panels (a-c) show the best fitting models,
optimized using a simplex algorithm, and RMSE values give the root mean
squared errors of the fits.}

@@

Following Meese et al. (2006), we also inspected the slope of the
psychometric function for each condition (see
Figure~\ref{fig-dipperfig}d-f). At detection threshold, slopes were
relatively steep for all stimuli, with values around \(\beta=4\). As
pedestal contrasts increased, slopes linearized and reduced to around
\(\beta=1.3\) (Foley \& Legge, 1981; Meese et al., 2006), and remained
shallow at high pedestal contrasts. The exception to this was the
dichoptic condition, where slopes became extremely steep at high
dichoptic mask contrasts, consistent with previous observations (Baker
et al., 2013; Meese et al., 2006). This was clear for all three data
sets, with slope values in the range \(4 < \beta < 8\). However, we
observe that slope estimates are more variable than threshold estimates
(note the large error bars), particularly when using adaptive
staircases, which deploy most trials close to threshold. Our second
experiment therefore investigated the slope of the psychometric function
in more detail for dichoptic masking using the method of constant
stimuli, as well as exploring dichoptic interactions between chromatic
channels.

@@

\caption{\label{fig-MCSfig}Summary of data from Experiment 2. Panels
(a-c) show psychometric functions for each condition, pooled across
participants (600 trials per target contrast level). Panels (d-f) show
threshold and slope estimates for individual participants (points) and
the boundary of the posterior density estimates for fits to the pooled
data (ellipses). Panel (g) shows the average threshold elevation factor
for each combination of target and mask stimulus. Panel (h) shows the
geometric mean psychometric slope value for each masking condition in
Weibull \(\beta\) units. Contrast values are expressed relative to the
target only threshold for each stimulus type.}

@@

Threshold elevation was greatest when the target and mask had the same
chromaticity - notice that the psychometric function is shifted furthest
to the right for the achromatic target with an achromatic mask (white
and black circles in Figure~\ref{fig-MCSfig}a), for the red/green target
with a red/green mask (red and green circles in
Figure~\ref{fig-MCSfig}b), and for the blue/yellow target with a
blue/yellow mask (blue and yellow circles in Figure~\ref{fig-MCSfig}c).
Masking was weakest between achromatic masks/targets and chromatic
masks/targets. Finally, there was an intermediate level of masking
between red/green and blue/yellow stimuli. This is summarised in
Figure~\ref{fig-MCSfig}g, which represents threshold elevation for each
combination of target and mask chromaticity. Note that the positive
diagonal exhibits the highest values, and represents threshold elevation
between targets and masks of the same chromaticity.
